---
title: "MATH 9102 - Probability and Statistical Inference Assignment - 4"
author: "Antonio Silva (D23129331@mytudublin.ie)"
output:
  html_document:
    toc: true
    toc_depth: 2
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
  pdf_document:
    toc: true
    toc_depth: '2'
---

```{r, echo = FALSE, results = "hide"}
needed_packages <- c("stats", "lm.beta", "stargazer", "ggplot2", "ppcor", "car", "nortest", "dplyr", "tidyverse")                      
# Extract not installed packages
not_installed <- needed_packages[!(needed_packages %in% installed.packages()[ , "Package"])]    

if(length(not_installed)) install.packages(not_installed, repos ="http://cran.us.r-project.org")                              
library(stats, quietly = T)
library(ggplot2, quietly = T)
library(lm.beta, quietly = T)
library(stargazer, quietly = T)
library(ppcor, quietly = T)
library(car, quietly = T)
library(nortest, quietly = T)
library(dplyr, quietly = T)
```

# Question 1 - Linear regression with dummy variable [5 marks]
Consider the dataset "weatherhistory.csv".

 a) Describe dependent variable pressure and independent variable temperature.
 b) Explore the relationship between pressure and temperature.
 c) Build a linear model considering temperature and pressure.
 d) Identify a dummy variable and build extended model considering dummy variable.
 e) Report your findings.

## Answer
### a) Describe dependent variable pressure and independent variable temperature.
Load the dataset

```{r, echo=TRUE}
weatherHistory <- read.csv("weatherHistory.csv")
str(weatherHistory)
```

```{r, echo=TRUE}
head(weatherHistory)
```

```{r, echo=TRUE}
colnames(weatherHistory)
```
```{r, echo=TRUE}
ggplot(weatherHistory, aes(y = temperature)) +
  geom_boxplot(fill = "lightblue", colour = "darkblue") +
  labs(title = "Boxplot of Temperature", y = "Temperature") +
  theme_minimal()
```

```{r, echo=TRUE}
ggplot(weatherHistory, aes(y = pressure)) +
  geom_boxplot(fill = "lightblue", colour = "darkblue") +
  labs(title = "Boxplot of Temperature", y = "Temperature") +
  theme_minimal()
```

```{r, echo=TRUE}
ggplot(weatherHistory, aes(x = temperature)) +
  geom_histogram(bins = 60, fill = "skyblue", color = "black") + # You can adjust the number of bins
  labs(title = "Histogram of Temperature", x = "Temperature (°C)", y = "Frequency") +
  theme_minimal()
```

```{r, echo=TRUE}
ggplot(weatherHistory, aes(x = pressure)) +
  geom_histogram(bins = 200, fill = "lightgreen", color = "black") + # Adjust the number of bins as needed
  labs(title = "Histogram of Pressure", x = "Pressure (hPa)", y = "Frequency") +
  theme_minimal()
```

```{r, echo=TRUE}
psych::describe(weatherHistory[, c("temperature", "pressure")])
```

Temperature has the mean 11.93 and the a standard deviation of 9.55. With this
standard deviation showing significant fluctuations in the temperature.
With the kurtosis of -0.57 indicates a distribution flatter than a normal 
distribution. We also have 0.03 of standard error of the mean that the measure 
is precise.
We can conclude that temperature is a symmetric and a flatter variable.

Pressure data in the other hand shows a skew data with a lot of zero values.
Hit has a kurtosis of 69.26 indicate thick tails and a skew of -8.42 indicating
the data is concentrated in the right side of the mean. 


Testing the temperature for normality

```{r, echo=TRUE}
ggplot(weatherHistory, aes(sample=temperature)) + stat_qq()
```

```{r, echo=TRUE}
ad.test(weatherHistory$temperature)

ad.test(weatherHistory$pressure)
```

We can also see that neither temperature or pressure follows a normal 
distribution.

### b) Explore the relationship between pressure and temperature.

To explore the relationship between pressure and temperature.

```{r, echo=TRUE}
ggplot(weatherHistory, aes(x = temperature, y = pressure)) +
  geom_point(alpha = 0.5) +  # using some transparency for points
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +  
  labs(title = "Temperature vs. Pressure",
       x = "Temperature",
       y = "Pressure") +
  theme_minimal()
```
It seems that the pressure outliers are affecting the relationship between 
temperature and pressure. Let's remove them e see the correlation.

```{r, echo=TRUE}
weatherHistoryWithPressure <- weatherHistory %>%  
    filter(pressure != 0) %>%  
    filter(!is.na(temperature) & !is.na(pressure))


ggplot(weatherHistoryWithPressure, aes(x = temperature, y = pressure)) +
  geom_point(alpha = 0.5) +  # using some transparency for points
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) + 
  labs(title = "Temperature vs. Pressure",
       x = "Temperature",
       y = "Pressure") +
  theme_minimal()
```

Let's also do a Pearson's correlation between the variables.

```{r, echo=TRUE}
cor.test(weatherHistoryWithPressure$temperature, weatherHistoryWithPressure$pressure)
```
The scatterplot suggest that we have a non-linear relationship between 
temperature and pressure. So doing a linear regression model will not capture
or explain the relationship between them.

The correlation coefficient from the Pearson's test indicates a negative 
relationship of 0.31. So it means when the temperature increases the pressure
decreases.

The p-value is also too low (2.2e-16) suggesting the relationship is statistical
significant.

### c) Build a linear model considering temperature and pressure.

First let's create the model.

```{r, echo=TRUE}
lmTemperaturePressure <-  lm(pressure ~ temperature, 
                             data = weatherHistoryWithPressure)
anova(lmTemperaturePressure)
```

```{r, echo=TRUE}
summary(lmTemperaturePressure)
```

Standartize co-efficients to get the result expressed in standart deviations.
```{r, echo=TRUE}
lm.beta(lmTemperaturePressure)  
```

```{r, echo=TRUE}
suppressWarnings(stargazer(lmTemperaturePressure, type="text"))
```
So we got the following results:

- For every 1 degree increase in temperature the pressure decreases 0.253 units.
- The p value is 0.01 who give us 99% confidence level. Also our F statistic
is 10,150.310. Suggesting that the temperature is a good predictor of pressure.
- R squared is 0.096 meaning that 9.6% of the pressure variance is explained
by the temperature.


```{r, echo=TRUE}
plot(residuals(lmTemperaturePressure), main = "Residuals of the Model",
     xlab = "Index", ylab = "Residuals")
abline(h = 0, col = "red")
```


```{r, echo=TRUE}
plot(lmTemperaturePressure)
```

Looking at the residuals we can see that there is a funneling effect 
showing a spread out as the fitted values increase suggesting a non-constant
variance.

#### d) Identify a dummy variable and build extended model considering dummy variable.

Let's use the categorical variable precipType as a dummy variable for our model
 
```{r, echo=TRUE}
weatherHistoryDummy <- weatherHistory
weatherHistoryDummy$humidityScale <- cut(weatherHistory$windSpeed, 
                                     breaks=c(0, 0.25, 0.6, 1), 
                                     labels=c("dry", "comfortable", "humid"))
levels(weatherHistoryDummy$humidityScale)
table(weatherHistoryDummy$humidityScale)
```
 
So we have we possible values: dry, rain and snow.

Mapping them in our dataset.

```{r, echo=TRUE}
weatherHistoryDummy$dry <- ifelse(weatherHistoryDummy$humidityScale == "dry", 1, 0)
weatherHistoryDummy$comfortable <- ifelse(weatherHistoryDummy$humidityScale == "comfortable", 1, 0)
weatherHistoryDummy$humid <- ifelse(weatherHistoryDummy$humidityScale == "humid", 1, 0)

table(weatherHistoryDummy[, c("humidityScale")])
table(weatherHistoryDummy[, c("dry")])
table(weatherHistoryDummy[, c("comfortable")])
table(weatherHistoryDummy[, c("humid")])
```

Lets consider the humid dummy variable and how it influenciates the pressure.

```{r, echo=TRUE}
lmTempPresWithDummy <- lm(pressure ~ temperature + humid, data = weatherHistoryDummy)
```

### e) Report your findings.
 
```{r, echo=TRUE}
summary(lmTempPresWithDummy)
```

```{r, echo=TRUE}
lm.beta(lmTempPresWithDummy)  
```

```{r, echo=TRUE}
suppressWarnings(stargazer(lmTempPresWithDummy, type="text"))
```

If we consider the humidity as dummy variable we get the model equaltion:
$$ pressure = 1010.012 - 0.085 \times temperature + 6.052 \times humid  $$

Conclusions: 
 
 - The pressure is 1010.012 when the temperature and humidity is 0;
 - The pressure drops 0.085 per degree increase in temperature;
 - When the weather is humid the pressure increases 6.052;
 - R-squared is less than 0.001 indicating that the model explain less than 1%
   of the variance to the pressure;
 - A negative adjusted R-squared suggests that the model fits the data worse than
   just the mean of the dependent variable. 
 -  F-statistic is 0.4302 with a p-value of 0.6505, indicating that the model is not statistically significant.
 